%\begin{refsection}

%\chapter{The search for allopatric speciation in the gut microbial communities of Tanganyikan cichlids}

\chapter{Prediction of ecological function in the microbiome using machine learning on the graph spectra of coevolving subnetworks}

\chapterauthor{Russell Y. Neches, Matthew D. McGee, Peter C. Wainwright, Jonathan A. Eisen}

\section{Author contributions}

This project was conceived, initiated and carried to completion by Russell Y. Neches with input from Peter C. Wainwright, in collaboration with Matthew D. McGee and in collaboration with, and with support and mentorship from Jonathan A. Eisen. With the exception of animal handling, experiments were planned and performed by Russell Y. Neches. Animals were handled by Matthew D. McGee under the supervision of Peter C. Wainwright. Experimental apparatus was constructed by Russell Y. Neches. Sample preparation and DNA sequencing were performed by Qingyi Zhang. Software was written and analysis was performed by Russell Y. Neches with advice from Matthew D. McGee. The manuscript was written by Russell Y. Neches.

\subfile{FishPoo/abstract}
\subfile{FishPoo/introduction}
\subfile{FishPoo/background}
\subfile{FishPoo/methods}
\subfile{FishPoo/discussion}
\subfile{FishPoo/conclusion}


\section{Funding}

RYN was funded by a grant from the Alfred P. Sloan Foundation to Jonathan A. Eisen. 

\printbibliography[heading=subbibliography]
%\printbibliography[segment=2]

\subfile{FishPoo/tables/table2}
\subfile{FishPoo/tables/table3}

%\end{refsection}